---
title: 'PS5: Markdown, Basic Plots'
author: "Michaelangelo Labrador"
date: "2023-02-14"
output: html_document
---

## Load and check data

### 1. Created rmarkdown document called assignment.rmd
```{r}
getwd()
list.files()
```

### 2. Load data.

```{r message = FALSE}
library(readr)
library(tidyverse)
gapminder <- read_delim("gapminder.csv.bz2")
```

The _gapminder_ data has `r nrow(gapminder)` rows and `r ncol(gapminder)` 
columns. 

### 3. Print a small sample of the data. 

```{r}
gapminder %>% 
  sample_n(5) %>% 
  knitr::kable()
```

Seems like there are a lot of NA values for the region column of the data set. 
Other than that, the data looks good. 

## Descriptive Statistics

### 2. How many countries in the data set?

```{r}
# iso3 col
iso_3 <- gapminder %>% 
  select(iso3) %>% 
  n_distinct() 
```

```{r}
# name col
names <- gapminder %>% 
  select(name) %>% 
  n_distinct(.) 
```
```{r}
iso_2 <- gapminder %>% 
  select (iso2) %>% 
  n_distinct() 
  
```

Iso3 has `r iso_3` countries. Name column has `r names` countries. Finally, 
iso2 has `r iso_2` countries. 

### (a) Find how many names are there for each iso-2 code. Are there any iso-2 
### codes that correspond to more than one name? What are these countries?

```{r}
gapminder %>% 
  group_by(iso2)%>% 
  summarize(n = n_distinct(name)) %>% 
  filter( n > 1) %>% 
  knitr::kable()
```  

```{r}

gapminder %>% 
  group_by(name) %>% 
  filter(is.na(iso2)) %>% 
  summarize(n = n_distinct(iso2)) %>% 
  knitr::kable()
```


There is 1 name for each iso-2 code. However, one iso-2 code has 2 names and that is
NA (missing value).

### (b) Now repeat the same for name and iso3-code. Are there country names 
### that have more than one iso3-code? What are these countries?

```{r}
gapminder %>% 
  group_by(name) %>%
  summarize(n = n_distinct(iso3)) %>% 
  filter(n > 1) %>% 
  knitr::kable()
```  
```{r}
gapminder %>% 
  group_by(name) %>% 
  distinct(iso3) %>% 
  filter(is.na(name)) %>%
  knitr::kable()
  
```  
The country NA(missing value) has 4 iso-3 codes.

### 3. What is the minimum and maximum year in these data?
```{r}
gapminder %>% 
  select(time) %>% 
  filter(!is.na(time)) %>% 
  summarize(min = min(time), max = max(time)) %>% 
  pull(min, max)
```

## CO2 Emissions

### 1. How many missing co2 emissions are there for each year? Anaylyze both
### missing CO2 and co2_pc. Which years have most missing data?

```{r}
# CO2 column
missing_co2 <- gapminder %>% 
  group_by(time) %>% 
  filter(is.na(co2)) %>% 
  count(co2) %>%
  head(10) %>% 
  knitr::kable() %>% 
  print()
```

```{r}
# co2_PC column 
missing_co2_pc <- gapminder %>% 
  group_by(time) %>% 
  filter(is.na(co2_PC)) %>% 
  count(co2_PC) %>% 
  head(10) %>% 
  knitr::kable() %>% 
  print()
```

The amount of missing data is the same for both  columns. Starts off in as 60 
missing values in 1960. Then trickles down to about 13 missing values in 2016.

```{r}
most_missing <- gapminder %>% 
  group_by(time) %>% 
  filter(is.na(co2_PC)) %>% 
  count(co2_PC) %>%
  arrange(desc(n)) %>% 
  head(3) %>% 
  pull(time) %>%
  print()
```

### 2. Make a plot of total CO2 emissions over time for U.S., China, and India. 
### Add a few countries of your choice. Explain what do you see. 

```{r}
co2_plot <- gapminder %>%
  group_by(time,name) %>% 
  filter(name == "India"|
           name == "China"| 
           name == "United States of America"|
           name == "Canada" |
           name == "Aruba", 
         !is.na(co2))

  ggplot(data = co2_plot,
         aes(x = time, y = co2,
             group = name,
             color = name)) + 
  geom_line(linewidth = 0.5) +
    labs(title = "Total CO2 Over time",
         x = "Time",
         y = "Total CO2")
  
```

It looks like China had a huge boost of total co2 in 2000 and exponentially rose.
United Sates takes 2nd place but has a more constant co2 levels. Other countries
such as Aruba, Canada, and India have much lower co2 levels. Could be based on 
their lower co2 production based on manufacturing and other societal practices. 

### 3. Now let's analyze the CO2 emissions per capita (co2_PC). Make a similar plot 
### of the same countries. What does this figure suggest?
```{r}
co2_PC_plot <- gapminder %>%
  group_by(time,name) %>% 
  filter(name == "India"|
           name == "China"| 
           name == "United States of America"|
           name == "Canada" |
           name == "Aruba", 
         !is.na(co2))

  ggplot(data = co2_PC_plot,
         aes(x = time, y = co2_PC,
             group = name,
             color = name)) + 
  geom_line(linewidth = 0.5) +
    labs(title = "Total CO2 Per Capita Over time",
         x = "Time",
         y = "Total CO2")
  
```


All co2 levels per capita are nearly the same an constant except for Aruba but
it takes a nose dive to equal out to similar levels as other countries. Figure
suggests that this is taking into account based on population. Although co2
levels are rising, population keeps it at a low, constant average. 

### 4. Compute average CO2 emissions per capita across the continents (assume
### region is the same as continent). Comment what do you see. 

```{r}
 gapminder %>% 
  group_by(region,time) %>% 
  filter(!is.na(co2_PC), !is.na(name), time == 2016) %>% 
  summarize(mean_co2_PC = mean(co2_PC)) %>% 
  knitr::kable()
```


Africa has the lowest co2 per capita. Americas comes next. And the two highest
are Europe and Oceania. 

### 5. Make a barplot where you show the previous results-average CO2 emissions
### per capita across continents in 1960 and 2016. 

```{r}


gapminder %>% 
  group_by(region, time) %>% 
  filter(!is.na(co2_PC), !is.na(name), time == 2016 | time == 1960) %>% 
  reframe(mean = mean(co2_PC)) %>% 
  ggplot(aes(x = region, y = mean, fill = as.factor(time))) + 
    geom_bar(stat = "identity", position = "dodge") + 
  labs(x = "continent",
       y = "Average CO2 per capita",
       fill = "Year")
  

```

### 6. Which countries are the three largest, and three smallest CO2 emiitters
### (in terms of CO2 per capita) in 2019 for each continent? (Used 2016 instead)

```{r}
gapminder %>% 
  group_by(region) %>% 
   filter(time == 2016, !is.na(co2_PC), !is.na(region), !is.na(name)) %>% 
  filter(rank(desc(co2_PC)) <= 3) %>% 
  arrange(region) %>% 
  select(region, name, co2_PC) %>% 
  knitr::kable()
```
The _top 3_ emitters for each continent are:

- Africa - Libya, Seychelles, and South Africa
- Americas - Canada, Trinidad and Tobago, U.S.A.
- Asia - Bahrain, Kuwait, Qatar
- Europe - Faroe Islands, Gibraltar, Luxembourg
- Oceania - Australia, New caledonia, Palau

```{r}
gapminder %>% 
  group_by(region) %>% 
   filter(time == 2016, !is.na(co2_PC), !is.na(region), !is.na(name)) %>% 
  filter(rank(co2_PC) <= 3)  %>% 
  arrange(region) %>% 
  select(region, name, co2_PC) %>% 
  knitr::kable()
```

The 3 lowest co2 emitters per capita are:

- Africa - Burundi, Congo, Somalia
- Americas - Honduras, Haiti, Nicaragua
- Asia - Afghanistan, Nepal, Yemen
- Europe - Albania, Liechtenstein, Moldova
- Oceania - Kiribati, Solomon Islands, Vanuatu




## GPD per Capita

### 1. Make a scatterplot of GPD per capita versus life expectancy by country, 
### using data from 1960. Make the point size dependent on the country size, 
### and color those according according to the continent. Feel free to adjust 
### the plot in other ways to make it better. Comment what you see.

```{r}
library(tidyverse)
gapminder %>% 
  group_by(name) %>% 
  filter(!is.na(lifeExpectancy), !is.na(GDP_PC), time == 1960) %>% 
  ggplot(aes(x = GDP_PC, y = lifeExpectancy, color = region)) +
  geom_point(aes(size = totalPopulation)) + 
  labs( x = "GPD per capita",
        y = "Life Expectancy",
        size = "Total Population", 
        color = "Country by Region",
        title = "GDP per Capita by Country 1960")
```

Countries with low GDP per capita have a lower life expectancy. These countries
include mostly of Asia and Africa. Countries with with higher GDP per capita
have a higher life expectancy. This includes countries from mostly Americas and
Europe with the exception of one Oceania country. 

### Make a similar plot, but this time use 2019 data only. 

```{r}
library(tidyverse)
gapminder %>% 
  group_by(name) %>% 
  filter(!is.na(region), !is.na(lifeExpectancy), !is.na(GDP_PC), time == 2019) %>% 
  ggplot(aes(x = GDP_PC, y = lifeExpectancy, color = region)) +
  geom_point(aes(size = totalPopulation)) + 
  labs( x = "GPD per capita",
        y = "Life Expectancy",
        size = "Total Population", 
        color = "Country by Region",
        title = "GDP per capita by Country 2019")
```

### 3. Compare these two plots and comment what do you see. How as the world 
### developed through the last 60 years?

Africa's life expectancy and GPD has somewhat stayed the same as well as Asia. 
Asia, however, has a higher life expectancy now. American generally stayed the
same. Europe has a well but it seems some countries increased in GDP per capita. 

### 4. Compute the average life expectancy for each continent in 1960 and 2019. 
### Do the results fit with what do you see on the figures?

```{r}
gapminder %>% 
  group_by(region, time) %>% 
  filter(!is.na(lifeExpectancy), time == 1960 | time == 2019) %>% 
  summarize(avg_LE = mean(lifeExpectancy)) %>% 
  knitr::kable()
  
```

There is a definite growth of average of life expectancy for all continents. In
my comments, I said it looked like Africa didn't grow in life expectancy.
Further looking and comparing the two, the scale is much different. Although 
it looks the same, Africa did in fact grow in life expectancy. 

### Compute the average LE growth from 1960 and 2019 across the continents. Show
### the results in order of growth. Explain what do you see. 

```{r}
 gapminder %>% 
  group_by(time, region) %>% 
  filter(!is.na(time),
         !is.na(region), !is.na(lifeExpectancy), time == 1960 | time == 2019) %>% 
  reframe(avg_ov_time = mean(lifeExpectancy)) %>% 
  group_by(region) %>% 
  reframe(avg_growth_le = avg_ov_time - lag(avg_ov_time)) %>% 
  filter(!is.na(avg_growth_le)) %>% 
  arrange(avg_growth_le)


```

Africa and Asia had the highest growth in average life expectancy. Seems
as though the more underdeveloped countries caught up with standards of
Americas and Oceania which also had a growth but not as significant.
Europe had the lowest growth average but based on data we've already looked at, 
they have already had a pretty high life expectancy. 

### 6. Show the histogram of GPD per capita for years of 1960 and 2019. Try
### to put both histograms on the same graph, see how well you can do it!

```{r}
gapminder %>% 
  filter(!is.na(GDP_PC), !is.na(GDP_PC), time == 1960 | time == 2019) %>% 
  ggplot(aes(x = GDP_PC, fill = factor(time))) +
  geom_histogram(bins = 30, position = "dodge") +
  labs(fill = "Year")
```

### What was the ranking of US in terms of life expectancy in 1960 and in 2019?
### (When counting from the top.)

```{r}
gapminder %>% 
  filter(!is.na(lifeExpectancy), 
         !is.na(name), 
         time == 1960) %>% 
  mutate(rank = rank(desc(lifeExpectancy))) %>% 
  filter(name == "United States of America") %>% 
  select(name, rank) %>% 
  knitr::kable()
```
```{r}
 gapminder %>% 
  filter(!is.na(lifeExpectancy), 
         !is.na(name), 
         time == 2019) %>% 
  mutate(rank = rank(desc(lifeExpectancy))) %>% 
  filter(name == "United States of America") %>% 
  select(name, rank) %>% 
  knitr::kable()
```


### 8. What about the relative rank divided by the corresponding number of 
countries that have LE data in the corresponding year?

```{r}
gapminder %>% 
  filter(!is.na(lifeExpectancy), 
         !is.na(name), 
         time == 1960) %>% 
  mutate(rank = rank(desc(lifeExpectancy))) %>% 
  mutate(num_countries = n_distinct(name)) %>% 
  filter(name == "United States of America") %>% 
  summarize( rel_rank = rank/num_countries) %>%
  knitr::kable()
  
```
```{r}
gapminder %>% 
  filter(!is.na(lifeExpectancy), 
         !is.na(name), 
         time == 2019) %>% 
  mutate(rank = rank(desc(lifeExpectancy))) %>% 
  mutate(num_countries = n_distinct(name)) %>% 
  filter(name == "United States of America") %>% 
  summarize(rel_rank = rank/num_countries) %>% 
  knitr::kable()
  
```

### How long did you spent on this assignment?

**I lost track but I think I spent about anywhere from 6-10 hours on this 
assignment**
